
\documentclass{article}
\include{preamble.sty}

\begin{document}

  \section*{\sc Assignment: Article Review}

    Your task is to pick a research article published in an academic journal, read it, and write a summary.
  
    \subsection*{\sc Why are you being asked to do this?}

      Reading and summarizing research articles is a valuable skill for learning about new methods and preparing to do research in sport analytics. This assignment will give you practice in this skill.

    \subsection*{\sc What (exactly) are you being asked to do?}

      There are many different ways you can find your research article. Two academic journals you could explore are the {\it Journal of Quantitative Analysis in Sports} and the {\it Journal of Sports Analytics}. You could also enter search terms of interest into Google Scholar. If you need help finding your research article, let me know.

      In your summary (1-2 pages), please answer the following questions, roughly one paragraph per answer (labelling each section with the question it is answering).
      \begin{itemize}
        \item What did the authors do in this research?
        \item What is your favorite thing about this research?
        \item What is one remaining question you have about the research?
        \item What is one idea for future research that this article inspired for you?
        \item On a scale from 1 star to 5 stars, how would you rate this research article, and why?
      \end{itemize}

      \subsubsection*{\sc Submission Requirements}

        \begin{itemize}
          \item A PDF report (max 2 pages) answering the questions above
        \end{itemize}

    \subsection*{\sc How will your grade be determined?}
    
      This assignment will be graded for completion, meaning that as long as you successfully complete the assignment, your grade will be 100\%.

\end{document}