
\documentclass{article}

\usepackage[margin = 1in]{geometry}

\begin{document}

  \section*{\sc Assignment: Presentation}

    Your assignment is to carefully read an academic journal article which features mathematical modeling (statistics, computer science, applied math, operations research, etc.) applied to data from sports. Create a slide deck, and give a 20-minute presentation to the rest of the class summarizing the paper. Assume your audience has not read the paper.
    
    Please begin and end your presentation with a slide that answers the following questions:
    \begin{itemize}
      \item What was the sports research question the authors set out to answer?
      \item What was/were the mathematical model(s) the authors used?
      \item What was the novel contribution of this research paper?
    \end{itemize}

    \subsection*{\sc Selecting Your Paper}
      
      I will provide a list of suggested journal articles, but you are not required to select from this list. I encourage you to search Google Scholar for sports and modeling techniques that interest you. During this search, keep an eye out for papers in the following journals:

      \begin{itemize}
        \item The most preferred journals are statistics-discipline journals such as {\it Journal of the American Statistical Association}, {\it Journal of the Royal Statistical Society Series C}, {\it Annals of Applied Statistics} or {\it The American Statistician}.
        \item The next-most preferred journal is the {\it Journal of Quantitative Analysis in Sports}.
        \item Another journal you could consider is the {\it Journal of Sports Analytics}.
        \item There are no hard rules regarding the journal from which you select your paper. Some papers that do not appear in the above journals are better than some papers that do appear in the above journals.
      \end{itemize}

      When you think you have found your paper, please skim through it to make sure you are not biting off more than you can chew. The deadline to submit your selection is 8 days before your presentation.

\end{document}